---
title: "annotables"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    cache = FALSE,
    echo = TRUE,
    message = FALSE,
    warning = FALSE)
library(annotables)
```

[![DOI](https://zenodo.org/badge/3882/stephenturner/annotables.svg)](https://zenodo.org/badge/latestdoi/3882/stephenturner/annotables)

Provides tables for converting and annotating Ensembl Gene IDs.

## Installation

```{r, eval=FALSE}
install.packages("devtools")
devtools::install_github("stephenturner/annotables")
```

## Rationale

Many bioinformatics tasks require converting gene identifiers from one convention to another, or annotating gene identifiers with gene symbol, description, position, etc. Sure, [biomaRt](https://bioconductor.org/packages/release/bioc/html/biomaRt.html) does this for you, but I got tired of remembering biomaRt syntax and hammering Ensembl's servers every time I needed to do this.

This package has basic annotation information from **`r ensembl_version`** for the
species below. Gene counts are computed from the bundled tables; assemblies,
accessions, release dates and taxonomy IDs come from the [Ensembl REST
API](https://rest.ensembl.org).

```{r species_table, echo=FALSE, results='asis', message=FALSE}
library(dplyr)

# Assembly metadata for the assemblies these tables were built against.
# Retrieved from rest.ensembl.org (info/assembly, info/species, info/genomes).
meta <- tibble::tribble(
    ~dataset,      ~species,    ~taxon, ~assembly,            ~accession,         ~released,
    "grch38",      "Human",     9606L,  "GRCh38",             "GCA_000001405.29", "2013-12",
    "grch37",      "Human",     9606L,  "GRCh37",             "GCA_000001405.14", "2009-02",
    "grcm38",      "Mouse",     10090L, "GRCm38",             "GCA_000001635.8",  "2012-01",
    "rnor6",       "Rat",       10116L, "Rnor_6.0",           "GCA_000001895.4",  "2014-07",
    "galgal5",     "Chicken",   9031L,  "Gallus_gallus-5.0",  "GCA_000002315.3",  "2015-12",
    "wbcel235",    "Worm",      6239L,  "WBcel235",           "GCA_000002985.3",  "2012-12",
    "bdgp6",       "Fly",       7227L,  "BDGP6",              "GCA_000001215.4",  NA_character_,
    "drerio",      "Zebrafish", 7955L,  "GRCz11",             "GCA_000002035.4",  "2017-05",
    "mmul801",     "Macaque",   9544L,  "Mmul_8.0.1",         "GCA_000772875.3",  "2015-11",
    "cfamiliaris", "Dog",       9615L,  "CanFam3.1",          "GCA_000002285.2",  "2011-11",
    "sscrofa",     "Pig",       9823L,  "Sscrofa11.1",        "GCA_000003025.6",  "2017-02"
)

counts <- lapply(meta$dataset, function(n) {
    d <- get(n)
    tibble::tibble(
        dataset   = n,
        genes     = nrow(d),
        coding    = sum(d$biotype == "protein_coding", na.rm = TRUE),
        noncoding = sum(d$biotype != "protein_coding", na.rm = TRUE),
        tx2gene   = nrow(get(paste0(n, "_tx2gene")))
    )
}) %>% bind_rows()

meta %>%
    inner_join(counts, by = "dataset") %>%
    mutate(
        dataset  = paste0("`", dataset, "`"),
        released = ifelse(is.na(released), "—", released),
        across(c(genes, coding, noncoding, tx2gene), ~ format(.x, big.mark = ","))
    ) %>%
    knitr::kable(
        col.names = c("Dataset", "Species", "Taxon", "Assembly", "Accession",
                      "Assembly released", "Genes", "Coding", "Non-coding", "tx2gene"),
        align = "llrlllrrrr"
    )
```

Coding versus non-coding is `biotype == "protein_coding"` against everything
else; totals count gene rows, not unique symbols.

Note several of these assemblies have been superseded upstream (Ensembl 116
serves GRCm39, GRCr8, GRCg7b, ROS_Cfam_1.0 and Mmul_10). The dataset names
encode the assembly they were built from — see `NEWS.md` for the full mapping.

Where each table contains:

-   `ensgene`: Ensembl gene ID
-   `entrez`: Entrez gene ID
-   `symbol`: Gene symbol
-   `chr`: Chromosome
-   `start`: Start
-   `end`: End
-   `strand`: Strand
-   `biotype`: Protein coding, pseudogene, mitochondrial tRNA, etc.
-   `description`: Full gene name/description

Additionally, there are `tx2gene` tables that link Ensembl gene IDs to Ensembl transcript IDs.

## Usage

```{r, eval=FALSE}
library(annotables)
```

Look at the human genes table (note the description column gets cut off because the table becomes too wide to print nicely):

```{r}
grch38
```

Look at the human genes-to-transcripts table:

```{r}
grch38_tx2gene
```

Tables are saved in [tibble](http://tibble.tidyverse.org) format, pipe-able with [dplyr](http://dplyr.tidyverse.org):

```{r, results='asis'}
grch38 %>% 
    dplyr::filter(biotype == "protein_coding" & chr == "1") %>% 
    dplyr::select(ensgene, symbol, chr, start, end, description) %>% 
    head %>% 
    knitr::kable(.)
```

Example with [DESeq2](https://bioconductor.org/packages/release/bioc/html/DESeq2.html) results from the [airway](https://bioconductor.org/packages/release/data/experiment/html/airway.html) package:

```{r}
library(DESeq2)
library(airway)

data(airway)
airway <- DESeqDataSet(airway, design = ~cell + dex)
airway <- DESeq(airway)
res <- results(airway)

# tidy the results
res_tidy <- res %>%
    as.data.frame() %>%
    tibble::rownames_to_column("gene") %>%
    tibble::as_tibble() %>%
    dplyr::rename(estimate = log2FoldChange,
                  stderror = lfcSE,
                  statistic = stat,
                  p.value = pvalue,
                  p.adjusted = padj)
head(res_tidy)
```

```{r, results='asis'}
res_tidy %>% 
    dplyr::arrange(p.adjusted) %>% 
    head(20) %>% 
    dplyr::inner_join(grch38, by = c("gene" = "ensgene")) %>% 
    dplyr::select(gene, estimate, p.adjusted, symbol, description) %>% 
    knitr::kable(.)
```
