---
title: "annotables"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    cache = FALSE,
    echo = TRUE,
    message = FALSE,
    warning = FALSE)
library(annotables)
```

[![DOI](https://zenodo.org/badge/3882/stephenturner/annotables.svg)](https://zenodo.org/badge/latestdoi/3882/stephenturner/annotables)

Provides tables for converting and annotating Ensembl Gene IDs.

## Installation

```{r, eval=FALSE}
install.packages("pak")
pak::pkg_install("stephenturner/annotables")
```

## Rationale

Many bioinformatics tasks require converting gene identifiers from one convention to another, or annotating gene identifiers with gene symbol, description, position, etc. Sure, [biomaRt](https://bioconductor.org/packages/release/bioc/html/biomaRt.html) does this for you, but I got tired of remembering biomaRt syntax and hammering Ensembl's servers every time I needed to do this.

This package has basic annotation information from **`r ensembl_version`** for the
species below. Gene counts are computed from the bundled tables; assemblies,
accessions, release dates and taxonomy IDs come from the [Ensembl REST
API](https://rest.ensembl.org).

```{r species_table, echo=FALSE, results='asis', message=FALSE}
library(dplyr)

# Assembly metadata for the assemblies these tables were built against.
# Retrieved from rest.ensembl.org (info/assembly, info/species, info/genomes).
meta <- tibble::tribble(
    ~dataset,      ~species,    ~taxon, ~assembly,                     ~accession,         ~released,
    "grch38",     "Human",     9606L,  "GRCh38",                      "GCA_000001405.29", "2013-12",
    "grch37",     "Human",     9606L,  "GRCh37",                      "GCA_000001405.14", "2009-02",
    "grcm39",     "Mouse",     10090L, "GRCm39",                      "GCA_000001635.9",  "2020-06",
    "grcr8",      "Rat",       10116L, "GRCr8",                       "GCA_036323735.1",  "2024-01",
    "grcg7b",     "Chicken",   9031L,  "bGalGal1.mat.broiler.GRCg7b", "GCA_016699485.1",  "2021-01",
    "wbcel235",   "Worm",      6239L,  "WBcel235",                    "GCA_000002985.3",  "2012-12",
    "bdgp654",    "Fly",       7227L,  "BDGP6.54",                    "GCA_000001215.4",  NA_character_,
    "grcz11",     "Zebrafish", 7955L,  "GRCz11",                      "GCA_000002035.4",  "2017-05",
    "mmul10",     "Macaque",   9544L,  "Mmul_10",                     "GCA_003339765.3",  "2019-02",
    "roscfam1",   "Dog",       9615L,  "ROS_Cfam_1.0",                "GCA_014441545.1",  "2020-09",
    "sscrofa111", "Pig",       9823L,  "Sscrofa11.1",                 "GCA_000003025.6",  "2017-02"
)

counts <- lapply(meta$dataset, function(n) {
    d <- get(n)
    tibble::tibble(
        dataset   = n,
        genes     = nrow(d),
        coding    = sum(d$biotype == "protein_coding", na.rm = TRUE),
        noncoding = sum(d$biotype != "protein_coding", na.rm = TRUE),
        tx2gene   = nrow(get(paste0(n, "_tx2gene")))
    )
}) %>% bind_rows()

meta %>%
    inner_join(counts, by = "dataset") %>%
    mutate(
        dataset  = paste0("`", dataset, "`"),
        released = ifelse(is.na(released), "—", released),
        across(c(genes, coding, noncoding, tx2gene), ~ format(.x, big.mark = ","))
    ) %>%
    knitr::kable(
        col.names = c("Dataset", "Species", "Taxon", "Assembly", "Accession",
                      "Assembly released", "Genes", "Coding", "Non-coding", "tx2gene"),
        align = "llrlllrrrr"
    )
```

Coding versus non-coding is `biotype == "protein_coding"` against everything
else; totals count gene rows, not unique symbols.

### Dataset names

Datasets are named after the genome assembly. Recipes select by species, and
BioMart always returns whatever assembly is current for that species, so a
species-based name stops describing its own contents the moment Ensembl switches
assembly — silently, with nothing failing. Naming after the assembly makes a
stale name obvious.

Eight datasets were renamed in 0.3.0 for this reason. The old names were
removed, not deprecated, so existing code errors instead of quietly returning a
different assembly:

| Old name | New name | Species | Assembly |
|---|---|---|---|
| `rnor6` | `grcr8` | Rat | GRCr8 |
| `grcm38` | `grcm39` | Mouse | GRCm39 |
| `galgal5` | `grcg7b` | Chicken | bGalGal1.mat.broiler.GRCg7b |
| `mmul801` | `mmul10` | Macaque | Mmul_10 |
| `cfamiliaris` | `roscfam1` | Dog | ROS_Cfam_1.0 |
| `bdgp6` | `bdgp654` | Fruitfly | BDGP6.54 |
| `drerio` | `grcz11` | Zebrafish | GRCz11 |
| `sscrofa` | `sscrofa111` | Pig | Sscrofa11.1 |

`grch37`, `grch38` and `wbcel235` were already assembly-named and are unchanged.

Where each table contains:

-   `ensgene`: Ensembl gene ID
-   `entrez`: Entrez gene ID
-   `symbol`: Gene symbol
-   `chr`: Chromosome
-   `start`: Start
-   `end`: End
-   `strand`: Strand
-   `biotype`: Protein coding, pseudogene, mitochondrial tRNA, etc.
-   `description`: Full gene name/description

Additionally, there are `tx2gene` tables that link Ensembl gene IDs to Ensembl transcript IDs.

## Usage

```{r, eval=FALSE}
library(annotables)
```

Look at the human genes table (note the description column gets cut off because the table becomes too wide to print nicely):

```{r}
grch38
```

Look at the human genes-to-transcripts table:

```{r}
grch38_tx2gene
```

Tables are saved in [tibble](http://tibble.tidyverse.org) format, pipe-able with [dplyr](http://dplyr.tidyverse.org):

```{r, results='asis'}
grch38 %>%
    dplyr::filter(biotype == "protein_coding" & chr == "1") %>%
    dplyr::select(ensgene, symbol, chr, start, end, description) %>%
    head %>%
    knitr::kable(.)
```

Example with [DESeq2](https://bioconductor.org/packages/release/bioc/html/DESeq2.html) results from the [airway](https://bioconductor.org/packages/release/data/experiment/html/airway.html) package:

```{r}
library(DESeq2)
library(airway)

data(airway)
airway <- DESeqDataSet(airway, design = ~cell + dex)
airway <- DESeq(airway)
res <- results(airway)

# tidy the results
res_tidy <- res %>%
    as.data.frame() %>%
    tibble::rownames_to_column("gene") %>%
    tibble::as_tibble() %>%
    dplyr::rename(estimate = log2FoldChange,
                  stderror = lfcSE,
                  statistic = stat,
                  p.value = pvalue,
                  p.adjusted = padj)
head(res_tidy)
```

```{r, results='asis'}
res_tidy %>%
    dplyr::arrange(p.adjusted) %>%
    head(20) %>%
    dplyr::inner_join(grch38, by = c("gene" = "ensgene")) %>%
    dplyr::select(gene, estimate, p.adjusted, symbol, description) %>%
    knitr::kable(.)
```
